\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{proof-dashed,amsmath,amssymb,amsthm}
\usepackage{xcolor}
\usepackage{stmaryrd}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\input{program-macros}
\input{logic-macros}
\input{metatheory}

\title{Lecture 1: Natural Deduction}
\author{Chris Martens}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

Lecture outline:
\begin{itemize}
    \item Propositions, connectives, judgments
    \item Natural Deduction
    \item Proof terms
    \item Counting proofs
    \item Harmony
\end{itemize}

\section{Propositions, connectives, judgments}

Judgment: $\Gamma \vdash A \istrue$

\section{Inference rules for propositional natural deduction}

(Approximately Gentzen's NJ)

Conjunction:
\[
  \infer[\iand I]
  {\Gamma \vdash A \iand B \istrue}
  {\Gamma \vdash A \istrue
  &
  \Gamma \vdash B \istrue}
\qquad
  \infer[\iand E_1]
  {\Gamma \vdash A \istrue}
  {\Gamma \vdash A \iand B \istrue}
\qquad
  \infer[\iand E_2]
  {\Gamma \vdash B \istrue}
  {\Gamma \vdash A \iand B \istrue}
\]

Disjunction:
\[
  \infer[\ior I_1]
  {\Gamma \vdash A \ior B \istrue}
  {\Gamma \vdash A \istrue}
\qquad
  \infer[\ior I_2]
  {\Gamma \vdash A \ior B \istrue}
  {\Gamma \vdash B \istrue}
\]
\[
  \infer[\ior E]
  {\Gamma \vdash C \istrue}
  {\Gamma \vdash A \ior B \istrue
   &
   \Gamma, A\istrue \vdash C \istrue
   &
   \Gamma, B\istrue \vdash C \istrue
  }
\]

(The hypothetical judgment and hypothesis rule)

\[
  \infer[\mathsf{hyp}]
  {\Gamma \vdash A\istrue}
  {A\istrue \in \Gamma}
\]

Implication:

\[
  \infer[\imp I]
  {\Gamma \vdash A \imp B \istrue}
  {\Gamma, A\istrue \vdash B \istrue}
\qquad
  \infer[\imp E]
  {\Gamma \vdash B \istrue}
  {\Gamma \vdash A \imp B \istrue
  &
  \Gamma \vdash A \istrue}
\]

Truth and Falsehood:
\[
  \infer[\top I]
  {\top \istrue}{}
  \qquad
  \mathrm{(no\ \top E)}
  \qquad
  \mathrm{(no\ \bot I)}
  \qquad
  \infer[\bot E]
  {\Gamma \vdash C \istrue}
  {\Gamma \vdash \bot \istrue}
\]

Negation:

\[
\neg A := A \imp \bot
\]

\section{Examples}

Each of the following formulas $\phi$ can be shown 
$\cdot \vdash \phi \istrue$, i.e. they hold in the empty context, and can
be regarded as ``theorems'' of the system we have defined.

\begin{itemize}
  \item $(A \imp B) \iand A \imp B$
  \item $(A \imp (B \ior C)) \imp (A \iand \neg B) \imp C$
  \item $(A \ior B) \iand C \imp (A \iand C) \ior (B \iand C)$
\end{itemize}

\begin{exercise}\label{ex:derivs}
  Typeset derivations for these proofs.
\end{exercise}

\begin{exercise}
  \label{ex:hypnotation}
  In class, there was a discussion of the meaning of
  $\vdash$, the notation for the hypothetical judgment.
  Recall that we give the hypothetical judgment
  $B_1 \istrue, \ldots, B_n \istrue \vdash A \istrue$
  its meaning via ``scoped axioms'' of the form
  \[
    \infer[]{B_i \istrue}{}
  \]
  in the ambient metalogic, which are permitted to be
  used when proving $A \istrue$.

  In Gentzen's original notation, he elided the explicit mention of
  $\Gamma$, instead writing
  \[
  \infer[\imp I]
  {A \imp B \istrue}
  {\deduce{B \istrue}{[A \istrue]}}
  \]

  When we write $\Gamma$ explicitly, it is also very tempting
  to treat it as ``mere'' syntax that can be manipulated by
  inference rules (e.g. Weakening, Contraction, and Exchange; or an
  equational theory on contexts).

  These presentational choices impact both metatheory and
  pedagogy. For this exercise, decide for yourself on an approach that you
  might try using to present natural deduction to a audience unfamiliar
  with it: would you include $\Gamma$ explicitly or use Gentzen's (or a
  similar) notation? How would you write the hypothetical rule?
  Justify your choices.
  (If you like the presentation from lecture/these notes, that's fine, 
  but you still have to argue for it.)
  
\end{exercise}

\section{Soundness and Completeness}

What are some properties we believe this logic should have?

What stops us from defining the rules some other way?

What are the ``guiding principles''? What does it even mean to be a logic,
or a logical connective?

We won't be able to answer these questions in full yet, but we can at least
start with one proposed notion of soundness:
{\em It is not possible to derive $\cdot \vdash \bot \istrue$.}

How would we go about demonstrating such a fact?

\section{Counting Proofs}

Here is a related question: how many proofs are there of 

\[
  \cdot \vdash (A \imp B) \imp (A \iand B) \imp B \istrue
\]

?

There are at least two. However, there are in fact infinitely many.

\begin{exercise}
  Write two distinct proofs of 
  the above judgment.
\end{exercise}

\begin{exercise}
  Informally describe a procedure for generating infinitely many
  such proofs.
\end{exercise}


\section{Proof Terms}

For the remaining discussion, it will be convenient to have a more compact
notation for proofs. We therefore introduce {\em proof terms}.

We make a small change to three of our rules. First, for $\imp I$ and $\ior
E$, we add labels to the hypotheses:

\[
  \infer[\imp I^x]
  {\Gamma \vdash A \imp B \istrue}
  {\Gamma, x{:}A\istrue \vdash B \istrue}
\]

\[ 
\infer[\ior E^{x,y}]
  {\Gamma \vdash C \istrue}
  {\Gamma \vdash A \ior B \istrue
   &
   \Gamma, x{:}A\istrue \vdash C \istrue
   &
   \Gamma, y{:}B\istrue \vdash C \istrue
  }
\]

Then, we rename ``the'' hypothesis rule to its corresponding label:

\[
  \infer[x]
  {\Gamma \vdash A \istrue}
  {x{:}A \istrue \in \Gamma}
\]

These changes have the effect of allowing us to refer uniquely to
hypotheses that appear in the context $\Gamma$, and with them we can
abbreviate derivations of the preceding proofs.

In one more step, we apply the following syntactic translation:

\begin{tabular}{cc}
  $\imp I^x(M)$  & $\lambda{x}.\;M$ (function abstraction) \\
  $\imp E(M,N)$  & $M\; N$ (function application) \\
  $\iand I(M,N)$ & $(M, N)$ (pairs) \\
  $\iand E_i(M)$ & $\pi_i M$ (projection) \\
  $\ior I_i(M)$  & $\mathsf{in}_i\,M$ (injection)\\
  $\ior E^{x,y}(M,N,P)$ & $\excase{M, x.\,N, y.\,P}$ (case analysis)\\
  $\top I$ & $()$ (unit)\\
  $\bot E(M)$ & $\mathsf{abort}, M$ (error)
\end{tabular}

This gives us a somewhat familiar-looking syntax akin to functional
programming---more specifically, the simply-typed $\lambda$ calculus
(STLC).

\begin{exercise}
  Rewrite the derivations from Exercise \ref{ex:derivs} in proof term notation.
\end{exercise}

\begin{exercise}
  Revisit your decision from Exercise \ref{ex:hypnotation}.
  How do your choices affect the information recorded by the
  proof term and its relationship to the derivation?
  If necessary, design a modification to the proof term
  syntax to account for your decision.
\end{exercise}


% \section{Harmony: Local Soundness and Completeness}
% 
% \subsection{(Internal) Soundness and Completeness}
% 
% % (TODO additional discussion?)
% 
% {\bf Soundness:} the logic is not vacuous. That is,
% there are some propositions that are not true.
% 
% In particular, we can't write a closed proof of $\ifalse$.
% 
% {\bf Completeness:} the hypothetical judgment captures deduction in the logic.
% That is, even if we only allow the use of hypotheses
% at atomic propositions, then we can still construct
% a proof of each proposition by assuming it.
% 
% \subsection{Local Soundness as Proof Reduction}
% 
% Soundness essentially states that 
% a logic's proofs do not admit {\em more information}
% than what is used to construct them.
% The proof amounts to showing that {\em circuitous}
% steps in proofs can be eliminated. 
% Soundness is a {\em global} property of the logic:
% any rule could interact in some unexpected ways with
% all of the other rules, so we can't just check them in isolation.
% 
% However, there is a weaker notion of soundness that we can check for just a single
% propositional connective: if a proof has circuitousness by virtue of {\em immediately}
% introducing and then eliminating a connective, we can eliminate
% such a redundancy.
% %
% We demonstrate this by identifying ``circuitous'' proofs and showing
% how they can be rewritten to avoid the unnecessary steps.
% 
% Conjunction:
% 
% \[
% \infer[\iand{E}_1]
% {A \istrue}
% {
%     \infer[\iand{I}]
%     {A \iand B \istrue}
%     {
%         \deduce{A\istrue}{\DD_1}
%         &
%         \deduce{B\istrue}{\DD_2}
%     }
% }
% \qquad
% \Longrightarrow
% \qquad
% \deduce{A\istrue}{\DD_1}
% \]
% 
% \[
% \infer[\iand{E}_2]
% {B \istrue}
% {
%     \infer[\iand{I}]
%     {A \iand B \istrue}
%     {
%         \deduce{A\istrue}{\DD_1}
%         &
%         \deduce{B\istrue}{\DD_2}
%     }
% }
% \qquad
% \Longrightarrow
% \qquad
% \deduce{B\istrue}{\DD_2}
% \]
% 
% Implication:
% 
% \[
% \infer[\imp{E}]
% {B \istrue}
% {
% \infer[\imp{I}^u]
% {A \imp B \istrue}
% {   \deduce{B\istrue}
%     {\deduce{\vdots}{\infer[u]{A\istrue}{}}}
% }
% &
% \deduce{A}{\DD}
% }
% \qquad
% \Longrightarrow
% \qquad
% \deduce{B\istrue}
% {\deduce{\DD}{\infer[u]{A\istrue}{}}}
% \]
% 
% We will continue this discussion in the next lecture.
% 

\bibliographystyle{plainnat}
\bibliography{main}

\end{document}
